JWST Noise Floor I: Systematic Error Sources with JWST NIRCam Time Series

### 2020-05-02 ms.tex
small edits and reorganization mostly to pca and grism aperture sections

A standard long dark exposure used for detector characterization is a single integration that is 108 groups of 1 frame each (RAPID mode).
The detectors are read in full frame mode so that each frame is 10.73677 sec in duration with 4 output amplifiers in use, thus the total time for the exposure is 19.3 min.
We use one of these long dark integrations to create simulated time series as if it were observing many consecutive integrations in full frame mode.
The long darks are broken into 54 sequential pairs of images that are subtracted from each other (integration 1 minus integration 0, 3 minus 2, 5 minus 4 etc. in a zero-based counting scheme).
Figure \ref{fig:longDarkPhot} shows an example dark read pair from frame 55 minus frame 54.
In the raw data, there are 4 prominent vertical stripes that are 512 pixels wide and 2048 pixels tall due to offsets in the read-out amplifiers.
These amplifier offsets are caused by drifts in the reference voltage that are reset once per frame and vary from amplifier to amplifier.
Additionally, there are obvious correlations along the fast-read (X) direction due to 1/f noise that appear as horizontal stripes in the image.

@@

\subsubsection{PCA Correction}
Another approach to treating the 1/f noise is dimensionality reduction with Principal Component Analysis.
Here, we assume that the counts for each pixel along a column is a random variable that can be correlated with pixels in other columns.
Each row represents a different noise iteration.
First, we calculate the mean value of each pixel in the 108 group stack and subtract all groups by this mean image.
This removes all bias levels and focuses on the time variability of the pixels.
Here, we assume the dark current is negligible.
Next, we mask all bad pixels by setting all pixels with an absolute value larger than 200 DN and replace them with 0 DN to remove bad pixels that can drive the principal components.
The first 10 principal components are plotted in Figure \ref{fig:pcaEigenvectors} for two example groups (54 and 55).

The first 10 principal components explain 14\% of the total variance in each group up the ramp.
The principal components from one group to another are different, but the ordering of these components generally follows a pattern of increasing frequency with higher order principal components.
This is expected for 1/f noise where the lowest frequencies have most of the power.
Since the first principal components are different from one group to another, we calculated the principal components for each group independently.
We then create a noise model image which is the matrix multiplication of the eigenvectors by the principal components for 10 components.
This model is subtracted from each group to create a 1/f-corrected image.

The PCA subtraction is applied to each group up the ramp and then these groups are subtracted in pairs to create a time series.
We use 20 principal components as a starting point for this method.
The final time series has and measured standard deviation of 5,300 to 5,500 DN within the background-subtracted aperture.
This is comparable to the row and column subtraction standard deviation of 5,500 to 6,500 DN.
Adding 10 more principal components for a total of 20 eigenvectors only reduces the time series noise to 5,100 to 5,400 for the 3 apertures.

\begin{figure*}[!hbtp]
\centering
\includegraphics[width=.4\columnwidth]{pca_dark_amp_all_extra_bias_sub_grp_54.pdf}
\includegraphics[width=.4\columnwidth]{pca_dark_amp_all_extra_bias_sub_grp_55.pdf}
\caption{The first 10 principal component eigenvectors identified from group 54 (left) and group 55 (right).
The pre-amplifier reset values are seen in the first principal component, and change from one group (one read in RAPID) to another.
The principal components generally increase in frequency with principal component number.
In other words, the highest frequency principal components explain the least amount of the variance as would be expected for 1/f noise.
}\label{fig:pcaEigenvectors}
\end{figure*}

In another approach, we treat each amplifier separately so that the principal components are calculated independently.
This approach means that a noise signature in 3 amplifiers but not the fourth does not propagate to fourth amplifier when subtracting the PCA-based model.
This individual-amplifier essentially has four times the number of degrees of freedom, so we fit with only the first 5 principal components in each amplifier.
The resulting eigenvectors are shown in Figure \ref{fig:pcaEigenvectorsIndAmp}.

If we treat each amplifier separately, the PCA model is more effective in reducing 1/f noise.
We begin by keeping 5 principal components within each amplifier and subtracting these from each amplifier separately.
The resulting standard deviation of the time series is 2500 to 3820 DN.
This is improvement is beyond what is possible on a real source because the principal components include pixels over the source aperture.
To simulate a real photometry scenario, it is necessary to interpolate over the source aperture points as done in Section \ref{sec:smoothKernelSub} for the smoothing kernel.
However, we stop the PCA at this point because it does not do as well as the simple row-by-row amplifier-by-amplifier subtraction performed in Section \ref{sec:indAmpAvg}.

@@

\subsection{Grism Aperture}
The NIRCam grism time series are particularly susceptible to 1/f noise because the dispersion direction is parallel to the detector's fast-read direction.
This is visible in Figure \ref{fig:longDarkGrism} where the grism apertures (in red) are parallel to the 1/f noise correlations.
The background subtraction along the spatial direction (vertically) will efficiently remove pre-amplifier offsets from frame to frame and amplifier to amplifier.
However, there are no side reference pixels or horizontally-offset background regions available on the middle two amplifiers.
Therefore, the 1/f noise will adversely impact broadband behaviors of the lightcurves, especially on the middle two amplifiers.

We give some illustrative numbers for a 0.165 $\mu$m wide bin of a grism spectrum centered on 3.34 $\mu$m as listed in Table \ref{tab:noiseSummary1overf}.
Without reference pixel subtraction, the grism apertures have standard deviations of 77,600 DN.
Reference pixel subtraction drops this value to 9,640 DN and row-by-row subtraction drops it to 3,630 DN.
For a well depth of 60\%, there are $3.7 \times 10^6$ DN of source counts in the 0.17 $\mu$m bin, which amounts to a photon error of 1,420 DN assuming a gain of 1.84.
Therefore, 1/f noise can potentially dominate the error budget for broadband grism time series!
Expressed in ppm, the read noise after processing with row-by-row subtraction is 990 ppm as compared to a photon noise of 95 ppm for a CDS subtraction (2 groups in ramp).

@@

The timescale of most interest to studying atmospheres is the transit duration when trying to assess the stability.
The transit durations for most exoplanets considered is several hours long.
We therefore quantify error source over 1 hour duration.
If the smallest stripe subarray 2048$\times$64 is used, frame time is 0.34 seconds.
For simplicity, assuming a duty cycle of 80\% due to detector resets, there are 8460 frames over which to average read noise.
Therefore, the final read noise is only expected to be 1.4 ppm over a 1 hour timescale if each frame is statistically independent.
In section \ref{sec:darkBaseline}, we show from dark frames, that the read noise is statistically independent from integration to integration.

@@

%Simply variance weighting the spectra by the read and photon noise does not provide heavy enough weights.
%If we assume the read noise is perfectly correlated along the spectral direction and uncorrelated along the spatial direction, then equation \ref{eq:covarianceweights} becomes
%\begin{equation}
%w_j = \frac{P_j}{N_{ph} + N_{px} \sigma_R^2},
%\end{equation}
%where $N_{px}$ is the number of spectral pixels that will be binned and $\sigma_R$ is the read noise per pixel.
%Thus, for large values of $N_{px}$ (up to 512 for an amplifier), the $\sigma_R \approx 14e^-$ of read noise per pixel, can compound to 100,400 e$^-$ and the weighting approaches
%\begin{equation}
%w_j \propto P_j.
%\end{equation}
%
%Profile weighting should therefore lower the error on wavelength-binned time series.
%For the grism aperture experiment, we find that profile weighting gives a measured read noise of 355 ppm.
%Thus, profile weighting still does not reduce the standard deviation over an extraction aperture that is just 1 pixel wide at the peak grism flux.
%We show the real covariance matrix in Section \ref{sec:empCovarianceMatrix}.

@@

\begin{figure}[!hbtp]
\centering
\includegraphics[width=.6\columnwidth]{spec_spatial_cov.pdf}
\caption{The read noise shows significant correlations between pixels (unlike the traditional assumption that each pixel's noise is statistically independent).
The data was computed from a long dark exposure of 54 read pairs where reference pixel corrections and column-by-column and row-by-row subtractions have been performed.
The detector's diagonal elements are better than the read noise squared due to the row-by-row and column-by-column subtraction step.
Spatially, the pixels' have a covariance of 20~$(e^-)^2$ and a somewhat smaller spatial covariance of 10$(e^-)^2$.}\label{fig:covSpecSpatial}
\end{figure}

@@

%% See the notebooks wlp8_psf_image.ipynb and noise_sources_wlp8.ipynb for calculations
\subsection{The impact of correlated read noise on time series}
We consider here some example science impacts of correlated read noise in these photometric apertures.
We will evaluate the performance of a WEAK LENS +8 mode photometric time series, as will be common with NIRCam grism time series observations.
For a pessimistic scenario, we assume a 15,000 DN read noise contributes to every pair of reads for the 70 pixel radius aperture (and 75 to 100 pixel background subtraction annulus).
If the read noise is tolerable under this scenario, then it should have a minimal impact on photometric time series.
If it is a problem, we will discuss methods of mitigation such as the row-by-row subtraction above.

>>>>

A standard long dark exposure used for detector characterization is a single integration that is 108 groups of 1 frame each (RAPID mode).
The detectors are read in full frame mode so that each frame is 10.73677 sec in duration with 4 output amplifiers in use, thus the total time for the exposure is 19.3 min.
We use one of these long dark integrations to create simulated time series as if it were observing many consecutive integrations in full frame mode.
The long darks are broken into 54 sequential pairs of images that are subtracted from each other (integration 1 minus integration 0, 3 minus 2, 5 minus 4 etc. in a zero-based counting scheme).
We also re-label the times to include the reset time that would normally occur after each integration, so the

Figure \ref{fig:longDarkPhot} shows an example dark read pair from frame 55 minus frame 54.
In the raw data, there are 4 prominent vertical stripes that are 512 pixels wide and 2048 pixels tall due to offsets in the read-out amplifiers.
These amplifier offsets are caused by drifts in the reference voltage that are reset once per frame and vary from amplifier to amplifier.
Additionally, there are obvious correlations along the fast-read (X) direction due to 1/f noise that appear as horizontal stripes in the image.

@@

\subsubsection{PCA Correction}

\begin{figure*}[!hbtp]
\centering
\includegraphics[width=.4\columnwidth]{pca_dark_amp_all_extra_bias_sub_grp_54.pdf}
\includegraphics[width=.4\columnwidth]{pca_dark_amp_all_extra_bias_sub_grp_55.pdf}
\caption{The first 10 principal component eigenvectors identified from group 54 (left) and group 55 (right).
Pre-amplifier reset values are seen in the first principal component, and change from one group (one read in RAPID) to another.
These are residual values not fully subtracted by reference pixels.
The principal components generally increase in frequency with principal component number.
In other words, the highest frequency principal components explain the least amount of the variance as would be expected for 1/f noise.
}\label{fig:pcaEigenvectors}
\end{figure*}

Another approach to treating the 1/f noise is dimensionality reduction with Principal Component Analysis.
Here, we assume that the counts for each pixel along a column is a random variable that can be correlated with pixels in other columns.
Each row represents a different noise iteration.
First, we apply reference pixel correction to remove the pre-amplifier resets.
Next, we calculate the mean value of each pixel in the 108 group stack and subtract all groups by this mean image.
This removes all bias levels and focuses on the time variability of the pixels.
Here, we assume the dark current is negligible.
Next, we mask all bad pixels by setting all pixels with an absolute value larger than 200 DN and replace them with 0 DN to remove bad pixels that can drive the principal components.
The first 10 principal components are plotted in Figure \ref{fig:pcaEigenvectors} for two example groups (54 and 55).

The first 10 principal components explain 14\% of the total variance in each group up the ramp.
The principal components from one group to another are different, but the ordering of these components generally follows a pattern of increasing frequency with higher order principal components (smaller eigenvalues).
This is expected for 1/f noise where the lowest frequencies have most of the power.
Since the first principal components are different from one group to another, we calculated the principal components for each group independently.
We then create a noise model image which is the matrix multiplication of the eigenvectors by the principal components for 10 components.
This model is subtracted from each group to create a 1/f-corrected image.

The PCA subtraction is applied to each group up the ramp and then these groups are subtracted in pairs to create a time series.
We use 20 principal components as a starting point for this method.
The final time series has a measured standard deviation of 5,300 to 5,500 DN within the background-subtracted aperture.
This is comparable to the row and column subtraction standard deviation of 5,500 to 6,500 DN.
Adding 10 more principal components for a total of 20 eigenvectors only reduces the time series noise to 5,100 to 5,400 for the 3 apertures.

In another approach, we treat each amplifier separately so that the principal components are calculated independently.
This approach means that a noise signature in 3 amplifiers but not the fourth does not propagate to fourth amplifier when subtracting the PCA-based model.
This individual-amplifier model has four times the number of variables as the combined PCA approach, so we fit with only the first 5 principal components in each amplifier.
The resulting eigenvectors are shown in Figure \ref{fig:pcaEigenvectorsIndAmp}.

If we treat each amplifier separately, the PCA model is more effective in reducing 1/f noise.
We begin by keeping 5 principal components within each amplifier and subtracting these from each amplifier separately.
The resulting standard deviation of the time series is 2500 to 3820 DN.
This is improvement is beyond what is possible on a real source because the principal components include pixels over the source aperture.
To simulate a real photometry scenario, it is necessary to interpolate over the source aperture points as done in Section \ref{sec:smoothKernelSub} for the smoothing kernel.
However, we stop the PCA at this point because it does not do as well as the simple row-by-row amplifier-by-amplifier mean subtraction performed in Section \ref{sec:indAmpAvg}.

@@

\subsection{Grism Extraction}

The NIRCam grism time series are particularly susceptible to 1/f noise because the dispersion direction is parallel to the detector's fast-read direction.
This is visible in Figure \ref{fig:longDarkGrism} where the grism apertures (in red) are parallel to the 1/f noise correlations.
The background subtraction along the spatial direction (vertically) will efficiently remove pre-amplifier offsets from frame to frame and amplifier to amplifier.
However, there are no side reference pixels or horizontally-offset background regions available on the middle two amplifiers.
Therefore, the 1/f noise will adversely impact broadband behaviors of the lightcurves, especially on the middle two amplifiers for the F444W filter (3.9 to 5.0 $\mu$m over $\sim$2100 px) and right three amplifiers for the F322W2 filter (2.4 to 4.0 $\mu$m over $\sim$1600 px).

We give some illustrative numbers for a 0.165 $\mu$m wide bin of a grism spectrum centered on 3.34 $\mu$m as listed in Table \ref{tab:noiseSummary1overf}.
Without reference pixel subtraction, the grism apertures have standard deviations of 77,600 DN.
Reference pixel subtraction drops this value to 9,640 DN and row-by-row subtraction drops it to 3,630 DN.
For a well depth of 60\%, there are $3.7 \times 10^6$ DN of source counts in the 0.17 $\mu$m bin, which amounts to a photon error of 1,420 DN assuming a gain of 1.84.
Therefore, 1/f noise can potentially dominate the error budget for broadband grism time series.
Expressed in ppm, the read noise after processing with row-by-row subtraction is 990 ppm as compared to a photon noise of 95 ppm for a CDS subtraction (2 groups in ramp).

@@

@@

%Simply variance weighting the spectra by the read and photon noise does not provide heavy enough weights.
%If we assume the read noise is perfectly correlated along the spectral direction and uncorrelated along the spatial direction, then equation \ref{eq:covarianceweights} becomes
%\begin{equation}
%w_j = \frac{P_j}{N_{ph} + N_{px} \sigma_R^2},
%\end{equation}
%where $N_{px}$ is the number of spectral pixels that will be binned and $\sigma_R$ is the read noise per pixel.
%Thus, for large values of $N_{px}$ (up to 512 for an amplifier), the $\sigma_R \approx 14e^-$ of read noise per pixel, can compound to 100,400 e$^-$ and the weighting approaches
%\begin{equation}
%w_j \propto P_j.
%\end{equation}
%
%Profile weighting should therefore lower the error on wavelength-binned time series.
%For the grism aperture experiment, we find that profile weighting gives a measured read noise of 355 ppm.
%Thus, profile weighting still does not reduce the standard deviation over an extraction aperture that is just 1 pixel wide at the peak grism flux.
%We show the real covariance matrix in Section \ref{sec:empCovarianceMatrix}.
\begin{figure}[!hbtp]
\centering
\includegraphics[width=.6\columnwidth]{spec_spatial_cov.pdf}
\caption{The read noise shows significant correlations between pixels (unlike the traditional assumption that each pixel's noise is statistically independent).
The data was computed from a long dark exposure of 54 read pairs where reference pixel corrections and column-by-column and row-by-row subtractions have been performed.
The detector's diagonal elements are better than the read noise squared due to the row-by-row and column-by-column subtraction step.
Spatially, the pixels' have a covariance of 20~$(e^-)^2$ and a somewhat smaller spatial covariance of 10$(e^-)^2$.}\label{fig:covSpecSpatial}
\end{figure}

@@

@@

%% See the notebooks wlp8_psf_image.ipynb and noise_sources_wlp8.ipynb for calculations
\subsection{The impact of correlated read noise on time series}

The timescale of most interest to studying atmospheres is the transit duration when trying to assess the stability.
The transit durations for most exoplanets considered is several hours long.
We therefore quantify error source over 1 hour duration.
If the smallest stripe subarray 2048$\times$64 is used, frame time is 0.34 seconds.
For simplicity, assuming a duty cycle of 80\% due to detector resets, there are 8460 frames over which to average read noise.
Therefore, the final read noise is only expected to be 1.4 ppm over a 1 hour timescale if each frame is statistically independent.
%In section \ref{sec:darkBaseline}, we show from dark frames, that the read noise is statistically independent from integration to integration.

We consider here some example science impacts of correlated read noise in these photometric apertures.
We will evaluate the performance of a WEAK LENS +8 mode photometric time series, as will be common with NIRCam grism time series observations.
For a pessimistic scenario, we assume a 15,000 DN read noise contributes to every pair of reads for the 70 pixel radius aperture (and 75 to 100 pixel background subtraction annulus).
If the read noise is tolerable under this scenario, then it should have a minimal impact on photometric time series.
If it is a problem, we will discuss methods of mitigation such as the row-by-row subtraction above.
